---
title: "Survival Final Project"
author: "Xiaoluo Jiao"
date: "11/27/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(survival)
library(tidyverse)
library(StepReg)
library(Rcpp)
library(survminer)
library(MASS)

```

## Import dataset

```{r}
data(cancer, package = "survival")

cancer_df = lung
```

## Model

Firstly, construct the proportional hazards model we got in previous step:

```{r}
lung_df = cancer_df %>%
  mutate(status = case_when(status == 1 ~ 0,
                            status == 2 ~ 1)) %>%
  mutate(sex = as.factor(sex),
         ph.ecog = as.factor(ph.ecog), 
         status = as.factor(status))

lung_df_e = lung_df %>%
  mutate(ph.ecog.n = case_when(ph.ecog == 0 ~ 0,
                               ph.ecog == 1 ~ 1,
                               ph.ecog == 2 ~ 2,
                               ph.ecog == 3 ~ 2)) %>%
  mutate(ph.ecog.n = as.factor(ph.ecog.n))

lung_mod = lung_df_e %>%
  drop_na()

# model by stepwise
cox_mod_3 = coxph(Surv(time, status==1) ~ sex + ph.ecog.n + ph.karno + wt.loss, data = lung_mod)
```

# Diagnosis 

The Cox proportional hazards model makes two assumptions: \
(1) survival curves for different strata must have hazard functions that are proportional over the time t \
(2) the relationship between the log hazard and each covariate is linear, which can be verified with residual plots.

## Schoenfeld Residuals
```{r}
test.ph = cox.zph(cox_mod_3)
test.ph
```

The proportional hazard assumption is supported by a non-significant relationship between residuals and time, and refuted by a significant relationship. Here, with a significance level of 0.05, the test is statistically significant for `ph.karno`. The global test is not statistically significant, so our proportional hazards assumption is reasonable. 

```{r}
# residual plots for continuous covariates
ggcoxzph(test.ph, 
         var = c("ph.karno", "wt.loss"), 
         font.main = 12,
         point.col = "#F8766D"
         )
```

## Graphical Approach - log-log curves
### sex
```{r}
gplots <- list()

test_sex <- survfit(Surv(time, status == 1) ~ sex, data = lung_mod)

gplots[[1]] = ggsurvplot(test_sex, 
           fun = "cloglog",
           ggtheme = theme_bw(), 
           xlim = c(5, 1200),
           legend = "bottom", 
           legend.title = "Sex",
           legend.lab = c("Male", "Female"),
           xlab = "log(Time)",
           ylab = "log[-log(Survival Probability)]")
```

### ph.ecog
```{r}
test_ecog <- survfit(Surv(time, status == 1) ~ ph.ecog, data = lung_mod)

gplots[[2]] = ggsurvplot(test_ecog, 
           fun = "cloglog",
           ggtheme = theme_bw(),
           xlim = c(5, 1200),
           legend = "bottom", 
           legend.title = "ECOG Performance Score",
           legend.lab = c("0", "1", "2", "3"),
           xlab = "log(Time)",
           ylab = "log[-log(Survival Probability)]")
```

```{r}
# combine the log-log plots for categorical covariates
arrange_ggsurvplots(gplots, 
                    print = TRUE,
                    title = "Log of Negative Log of Estimated Survival Functions",
                    ncol = 2, 
                    nrow = 1)
```

### ph.karno
```{r}
data_phkarno = cancer_mod %>%
  mutate(ph.karno = case_when(ph.karno > 80 ~ "> 80",
                              ph.karno <= 80 ~ "<= 80")) 

test_phkarno <- survfit(Surv(time, status) ~ ph.karno, data = data_phkarno)

ggsurvplot(test_phkarno, 
           fun = "cloglog",
           ggtheme = theme_bw(), 
           legend = "bottom", 
           legend.title = "Karnofsky Score by Physician",
           legend.lab = c("<= 80", "> 80"))
```

### pat.karno
```{r}
data_patkarno = cancer_mod %>%
  mutate(pat.karno = case_when(pat.karno > 80 ~ "> 80",
                               pat.karno <= 80 ~ "<= 80")) 

test_patkarno <- survfit(Surv(time, status) ~ pat.karno, data = data_patkarno)

ggsurvplot(test_patkarno, 
           fun = "cloglog",
           ggtheme = theme_bw(), 
           legend = "bottom", 
           legend.title = "Karnofsky Score by Patient",
           legend.lab = c("<= 80", "> 80"))
```

### wt.loss
```{r}
avg_meal_cal = mean(as.numeric(cancer_df$meal.cal), na.rm = TRUE)

data_meal = cancer_mod %>%
  mutate(meal.cal = case_when(meal.cal > avg_meal_cal ~ "> average",
                              meal.cal <= avg_meal_cal ~ "<= average")) 

test_meal_cal = survfit(Surv(time, status) ~ meal.cal, data = data_meal)

ggsurvplot(test_meal_cal, 
           fun = "cloglog",
           ggtheme = theme_bw(), 
           legend = "bottom", 
           legend.title = "Calories Consumed at Meals",
           legend.lab = c("<= average", "> average"))
```

## Interaction for proportionality

Generate the time dependent covariates by creating interactions of the predictors and a function of survival time and include in the model.  If any of the time dependent covariates are significant then those predictors are not proportional.

```{r}
cox_mod_int = coxph(Surv(time, status == 2) ~ sex + ph.ecog.n + ph.karno + wt.loss +
                  sex*log(time) + ph.ecog.n*log(time) + ph.karno*log(time)  + wt.loss*log(time), 
                  cancer_df)

summary(cox_mod_int)
```

