---
title: "Survival Final Project"
author: "Xiaoluo Jiao"
date: "11/27/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(survival)
library(tidyverse)
library(StepReg)
library(Rcpp)
library(survminer)
library(MASS)

```

## Import dataset

```{r}
data(cancer, package = "survival")

cancer_df = lung
```

## Model

Firstly, construct the proportional hazards model we got in previous step:

```{r}
# drop missing values
cancer_mod = cancer_df %>%
 # mutate(status1 = ifelse(status==2,1,0)) %>% #???
  drop_na()

# model by stepwise
cox_mod_1 = coxph(Surv(time, status) ~ sex + ph.ecog + ph.karno + pat.karno + wt.loss, data = cancer_mod)
summary(cox_mod_1)
```

# Diagnosis 

The Cox proportional hazards model makes two assumptions: \
(1) survival curves for different strata must have hazard functions that are proportional over the time t \
(2) the relationship between the log hazard and each covariate is linear, which can be verified with residual plots.

## Schoenfeld Residuals
```{r}
test.ph = cox.zph(cox_mod_1)
test.ph
```

The proportional hazard assumption is supported by a non-significant relationship between residuals and time, and refuted by a significant relationship. Here, with a significance level of 0.05, the test is statistically significant for `ph.karno`. The global test is not statistically significant, so our proportional hazards assumption is reasonable. 

```{r}
ggcoxzph(test.ph)
```

## Graphical Approach - log-log curves
### sex
```{r}
test_sex <- survfit(Surv(time, status) ~ sex, data = cancer_mod)

ggsurvplot(test_sex, 
           fun = "cloglog",
           ggtheme = theme_bw(), 
           legend = "bottom", 
           legend.title = "Log-log Curve for Sex",
           legend.lab = c("Male", "Female"))
```

### ph.ecog
```{r}
data_ecog = cancer_mod %>%
  mutate(ph.ecog = as.factor(ph.ecog))

test_ecog <- survfit(Surv(time, status) ~ ph.ecog, data = data_ecog)

ggsurvplot(test_ecog, 
           fun = "cloglog",
           ggtheme = theme_bw(), 
           legend = "bottom", 
           legend.title = "Log-log Curve for ECOG Performance Score",
           legend.lab = c("0", "1", "2", "3"))
```

### ph.karno
```{r}
data_phkarno = cancer_mod %>%
  mutate(ph.karno = case_when(ph.karno > 80 ~ "> 80",
                              ph.karno <= 80 ~ "<= 80")) 

test_phkarno <- survfit(Surv(time, status) ~ ph.karno, data = data_phkarno)

ggsurvplot(test_phkarno, 
           fun = "cloglog",
           ggtheme = theme_bw(), 
           legend = "bottom", 
           legend.title = "Karnofsky Score by Physician",
           legend.lab = c("<= 80", "> 80"))
```

### pat.karno
```{r}
data_patkarno = cancer_mod %>%
  mutate(pat.karno = case_when(pat.karno > 80 ~ "> 80",
                               pat.karno <= 80 ~ "<= 80")) 

test_patkarno <- survfit(Surv(time, status) ~ pat.karno, data = data_patkarno)

ggsurvplot(test_patkarno, 
           fun = "cloglog",
           ggtheme = theme_bw(), 
           legend = "bottom", 
           legend.title = "Karnofsky Score by Patient",
           legend.lab = c("<= 80", "> 80"))
```

### wt.loss
```{r}
avg_meal_cal = mean(as.numeric(cancer_df$meal.cal), na.rm = TRUE)

data_meal = cancer_mod %>%
  mutate(meal.cal = case_when(meal.cal > avg_meal_cal ~ "> average",
                              meal.cal <= avg_meal_cal ~ "<= average")) 

test_meal_cal = survfit(Surv(time, status) ~ meal.cal, data = data_meal)

ggsurvplot(test_meal_cal, 
           fun = "cloglog",
           ggtheme = theme_bw(), 
           legend = "bottom", 
           legend.title = "Calories Consumed at Meals",
           legend.lab = c("<= average", "> average"))
```

