---
title: "Modeling"
author: "Yifei Xu"
date: "2022-11-24"
output: html_document
---

```{r setup, include=FALSE}
library(survival)
library(tidyverse)
library(survminer)
library(StepReg)
library(Rcpp)
```

```{r}
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 12, 
  fig.height = 20,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))
```


First, read in the dataset.

```{r, message=FALSE}
data(cancer, package="survival")

cancer_df=lung
```

Then, make comparison of the different groups of the attributes using the Kaplan-Meier Curve and conduct log-rank test.

### Sex
```{r, message=FALSE}
splots <- list()

# fit
cancer_sex = cancer_df %>%
  mutate(Sex = case_when(sex == 1 ~ "Male",
                         sex == 2 ~ "Female")) 

fit_sex = survfit(Surv(time, status) ~ Sex, data = cancer_sex) 
summary(fit_sex)$table

# visualization
splots[[1]] = ggsurvplot(fit_sex,
                         pval = TRUE, conf.int = TRUE,
                         surv.median.line = "hv", # Specify median survival
                         ggtheme = theme_bw(), # Change ggplot2 theme
                         legend = "bottom", 
                         legend.title = "Sex",
                         legend.lab = c("Male", "Female"))

# log-rank test
diff_sex = survdiff(Surv(time, status) ~ Sex, data = cancer_sex) 
diff_sex
```

### Age
```{r}
# fit
cancer_age = cancer_df %>%
  mutate(Age = case_when(age > 70 ~ "> 70",
                         age <= 70 ~ "<= 70")) 

fit_age = survfit(Surv(time, status) ~ Age, data = cancer_age)
summary(fit_age)$table

# visualization
splots[[2]] = ggsurvplot(fit_age,
                         pval = TRUE, conf.int = TRUE,
                         surv.median.line = "hv", 
                         ggtheme = theme_bw(), 
                         legend = "bottom", 
                         legend.title = "Age",
                         legend.lab = c("<= 70", "> 70"))

# log-rank test
diff_age = survdiff(Surv(time, status) ~ Age, data = cancer_age) 
diff_age
```

### ECOG performance score

```{r}
# fit
cancer_ecog = cancer_df %>%
  filter(!is.na(ph.ecog)) %>%
  mutate(ph.ecog = as.factor(ph.ecog))

fit_ecog = survfit(Surv(time, status) ~ ph.ecog, data = cancer_ecog)
summary(fit_ecog)$table

# visualization
splots[[3]] = ggsurvplot(fit_ecog,
                         pval = TRUE, conf.int = TRUE,
                         surv.median.line = "hv", 
                         ggtheme = theme_bw(), 
                         legend = "bottom", 
                         legend.title = "ECOG Performance Score",
                         legend.lab = c("0", "1", "2", "3"))

# log-rank test
diff_ecog = survdiff(Surv(time, status) ~ ph.ecog, data = cancer_ecog) 
diff_ecog
```

### Karnofsky performance score rated by physician
```{r}
# fit
cancer_phkarno = cancer_df %>%
  filter(!is.na(ph.karno)) %>%
  mutate(ph.karno = case_when(ph.karno > 80 ~ "> 80",
                         ph.karno <= 80 ~ "<= 80")) 

fit_phkarno = survfit(Surv(time, status) ~ ph.karno, data=cancer_phkarno)
summary(fit_phkarno)$table

# visualization
splots[[4]] = ggsurvplot(fit_phkarno,
                         pval = TRUE, conf.int = TRUE,
                         surv.median.line = "hv", 
                         ggtheme = theme_bw(), 
                         legend = "bottom", 
                         legend.title = "Karnofsky Score by Physician",
                         legend.lab = c("<= 80", "> 80"))

# log-rank test
diff_phkarno = survdiff(Surv(time, status) ~ ph.karno, data = cancer_phkarno) 
diff_phkarno
```

### Karnofsky performance score as rated by patient
```{r}
# fit
cancer_patkarno = cancer_df %>%
  filter(!is.na(pat.karno)) %>%
  mutate(pat.karno = case_when(pat.karno > 80 ~ "> 80",
                         pat.karno <= 80 ~ "<= 80")) 

fit_patkarno = survfit(Surv(time, status) ~ pat.karno, data=cancer_patkarno)
summary(fit_patkarno)$table

# visualization
splots[[5]] = ggsurvplot(fit_patkarno,
                         pval = TRUE, conf.int = TRUE,
                         surv.median.line = "hv", 
                         ggtheme = theme_bw(), 
                         legend = "bottom", 
                         legend.title = "Karnofsky Score by Patient",
                         legend.lab = c("<= 80", "> 80"))

# log-rank test
diff_patkarno = survdiff(Surv(time, status) ~ pat.karno, data = cancer_patkarno) 
diff_patkarno
```

### Calories consumed at meals
```{r}
# fit
avg_meal_cal = mean(as.numeric(cancer_df$meal.cal), na.rm = TRUE)

cancer_meal_cal = cancer_df %>%
  filter(!is.na(meal.cal)) %>%
  mutate(meal.cal = case_when(meal.cal > avg_meal_cal ~ "> average",
                         meal.cal <= avg_meal_cal ~ "<= average")) 

fit_meal_cal = survfit(Surv(time, status) ~ meal.cal, data = cancer_meal_cal)
summary(fit_meal_cal)$table


splots[[6]] = ggsurvplot(fit_meal_cal, 
                         pval = TRUE, conf.int = TRUE,
                         surv.median.line = "hv", 
                         ggtheme = theme_bw(), 
                         legend = "bottom",
                         legend.title = "Calories Consumed at Meals",
                         legend.lab = c("<= average", "> average"))

# log-rank test
diff_meal_cal = survdiff(Surv(time, status) ~ meal.cal, data = cancer_meal_cal) 
diff_meal_cal

```

### Weight loss

```{r}
# fit
avg_wt_loss = mean(as.numeric(cancer_df$wt.loss), na.rm = TRUE)

cancer_wt_loss = cancer_df %>%
  filter(!is.na(wt.loss)) %>%
  mutate(wt.loss = case_when(wt.loss > avg_wt_loss ~ "> average",
                         wt.loss <= avg_wt_loss ~ "<= average")) 

fit_wt_loss <- survfit(Surv(time, status) ~ wt.loss, data = cancer_wt_loss)
summary(fit_wt_loss)$table

# visualization
splots[[7]] = ggsurvplot(fit_wt_loss, 
                         pval = TRUE, conf.int = TRUE,
                         surv.median.line = "hv", 
                         ggtheme = theme_bw(), 
                         legend = "bottom",
                         legend.title = "wt_loss",
                         legend.lab = c("<= average", "> average"))

# log-rank test
diff_wt_loss = survdiff(Surv(time, status) ~ wt.loss, data = cancer_wt_loss) 
diff_wt_loss

```

```{r}
# combine the plots
arrange_ggsurvplots(splots, 
                    print = TRUE,
                    title = "Comparison of the different attribute groups using the Kaplan Meier Curve",
                    ncol = 2, 
                    nrow = 4)
```

Next, we fit a PH model.

```{r}
cancer_mod = cancer_df %>%
  mutate(status1 = ifelse(status==2,1,0)) %>% #???
  drop_na()

# full model
cox_mod_full = coxph(Surv(time, status) ~ sex + age + ph.ecog + ph.karno + pat.karno + meal.cal + wt.loss, data = cancer_mod)
summary(cox_mod_full)
```

Then we use stepwise method to select significant variables.

```{r, message=FALSE}
formula = Surv(time, status) ~ sex + age + ph.ecog + ph.karno + pat.karno + meal.cal + wt.loss

stepwiseCox(formula, 
            cancer_mod,
            include=NULL,
            selection=c("bidirection"),
            select="AIC",
            method=c("breslow"),
            sle=0.15,
            sls=0.15,
            weights=NULL,
            best=NULL)

cox_mod_1 = coxph(Surv(time, status) ~ sex + ph.ecog + ph.karno + pat.karno + wt.loss, data = cancer_mod)
summary(cox_mod_1)
```





