---
title: "Survival Group Project"
author: "Yujin Zhang"
date: '2022-11-20'
output: html_document
---


```{r setup, include=FALSE}
library(tidyverse)
library(survival)
library(ggplot2)
library(ggfortify)

knitr::opts_chunk$set(echo = TRUE)
```

# Load Data
```{r, message=FALSE, results='hide'}
lung
data(cancer, package="survival")

lung_df=lung %>% 
  mutate( sex = as.factor(sex),
          ph.ecog = as.factor(ph.ecog)) %>% 
  rename(ph_ecog = ph.ecog,
         ph_karno = ph.karno,
         pat_karno = pat.karno,
         meal_cal = meal.cal,
         wt_loss = wt.loss
         )

```
Dataset Sample:  
`r knitr::kable(head(lung_df, 7))`  

Dataset Attributes Overview:  
*inst:	    Institute Code  
*time:	    Survival time in days  
*status:	  Censoring status; 1= censored; 2 = dead
*sex:       Male=1; Female=2
*age:	      Age of the subject in years  
*ph_ecog:   Ecog performance score; 0: good, 5: dead  
*ph_karno:  Karnofsky performance score; 0: bad; 100: good rated by physician  
*pat_karno: Karnofsky performance score as rated by patient  
*meal_cal:	Calories consumed at meals  
*wt_loss:	  Weight loss in last six months 


# Overall Nelson-Aalen Estimator 
Nelson-Aalen hazard estimation focus on estimating the cumulative hazard function H(t)
```{r}
nelsonaalen = function(data, timevar, statusvar) {
  if (!is.data.frame(data)) {
    stop("Data must be a data frame")
  }
  timevar <- as.character(substitute(timevar))
  statusvar <- as.character(substitute(statusvar))
  time <- data[, timevar, drop = TRUE]
  status <- data[, statusvar, drop = TRUE]

  hazard <- survival::basehaz(survival::coxph(survival::Surv(time, status) ~ 1))
  idx <- match(time, hazard[, "time"])
  hazard[idx, "hazard"]
}

hazard = nelsonaalen(lung_df, time, status)
# plot(x = lung_df$time, main="Cumulative Probability for Event of Interest (Death)", y = hazard, ylab = "Cumulative Probability of subject's death", xlab = "Number of Days") 


lung_df1 = lung_df %>% 
  mutate(hazard = nelsonaalen(lung_df, time, status))

ggplot(data = lung_df1, aes(x = time, y = hazard)) + 
  geom_point(alpha = .3) +
  geom_line(colour = "blue") + 
  theme(legend.position = "bottom") +
  labs(
    title = "Cumulative Probability for Event of Interest (Death)",
    x = "Number of Days",
    y = "Cumulative Probability of subject's death"
  )
```

## Fleming-Harrington Estimator
Once the H(t) of Nelson-Aalen hazard estimator is obtained, Fleming-Harrington estimator for survival function can be obtained.
```{r}
survfit(Surv(time, status) ~ 1, lung_df, stype = 1, ctype = 2)
```
* Fleming-Harrington estimator for median survival is 310, which is larger/equal to K-M estimator?

# Nelson-Aalen Estimator on Variables of Interests
## Sex group
Dividing the population into two groups by the sex of the subject to check if the sex of a subject has any affect on the cumulative hazard.
```{r}
nelsonaalen = function(data, timevar, statusvar, invar) {
  if (!is.data.frame(data)) {
    stop("Data must be a data frame")
  }
  timevar <- as.character(substitute(timevar))
  statusvar <- as.character(substitute(statusvar))
  invar <- as.character(substitute(invar))
  time <- data[, timevar, drop = TRUE]
  status <- data[, statusvar, drop = TRUE]
  inv <- data[, invar, drop = TRUE]

  hazard <- survival::basehaz(survival::coxph(survival::Surv(time, status) ~ inv ))
  idx <- match(time, hazard[, "time"])
  hazard[idx, "hazard"]
}

lung_df1 = lung_df %>% 
  mutate(hazard = nelsonaalen(lung_df, time, status, sex),
         sexc = ifelse(sex != 1, "Female", "Male"))

ggplot(data = lung_df1, aes(x = time, y = hazard, color = sexc)) + 
  geom_point(alpha = .3) + 
  geom_line(colour = "blue") + 
  theme(legend.position = "bottom") +
  facet_grid(~sexc) +
  labs(title = "Cumulative Probability Adjusted for Sex", x = "Number of Days", y = "Cumulative Probability of subject's death")

```
Summary: From the graph it can be inferred that males have a comparatively lower chnaces of survival compared to females.

## Age group
Dividing the population into two groups (>=70 and < 70) by the ages of the subjects to check if the age of a subject has any affect on the survival.
```{r}
# add age group variable
lung_age =lung_df %>% 
  mutate(
    agegrp = ifelse(age >= 70, "Age >= 70", "Age < 70"),
    agegrpn = ifelse(age >= 70, 1, 0)
  ) %>% 
  mutate(agegrpn = as.factor(agegrpn))

# fit NA estimator
nelsonaalen = function(data, timevar, statusvar, invar) {
  if (!is.data.frame(data)) {
    stop("Data must be a data frame")
  }
  timevar <- as.character(substitute(timevar))
  statusvar <- as.character(substitute(statusvar))
  invar <- as.character(substitute(invar))
  time <- data[, timevar, drop = TRUE]
  status <- data[, statusvar, drop = TRUE]
  inv <- data[, invar, drop = TRUE]

  hazard <- survival::basehaz(survival::coxph(survival::Surv(time, status) ~ inv ))
  idx <- match(time, hazard[, "time"])
  hazard[idx, "hazard"]
}

lung_df1 = lung_age %>% 
  mutate(hazard = nelsonaalen(lung_age, time, status, agegrpn)
         )

ggplot(data = lung_df1, aes(x = time, y = hazard, color = agegrp)) + 
  geom_point(alpha = .3) + 
  geom_line(colour = "blue") + 
  theme(legend.position = "bottom") +
  facet_grid(~agegrp) +
  labs(title = "Cumulative Probability Adjusted for Age Group", x = "Number of Days", y = "Cumulative Probability of subject's death")

```
Summary: From the graph above, it can be inferred that subjects who are older than 70 years have a comparatively lower chances of survival compared to subjects younger than 70 years.

## Ecog performance score
```{r}
# fit NA estimator
nelsonaalen = function(data, timevar, statusvar, invar) {
  if (!is.data.frame(data)) {
    stop("Data must be a data frame")
  }
  timevar <- as.character(substitute(timevar))
  statusvar <- as.character(substitute(statusvar))
  invar <- as.character(substitute(invar))
  time <- data[, timevar, drop = TRUE]
  status <- data[, statusvar, drop = TRUE]
  inv <- data[, invar, drop = TRUE]

  hazard <- survival::basehaz(survival::coxph(survival::Surv(time, status) ~ inv ))
  idx <- match(time, hazard[, "time"])
  hazard[idx, "hazard"]
}

# drop missing ph_ecog obs
lung_df = lung_df %>% 
  drop_na(ph_ecog)

lung_df1 = lung_df %>% 
  mutate(hazard = nelsonaalen(lung_df, time, status, ph_ecog)
         )
ggplot(data = lung_df1, aes(x = time, y = hazard, color = ph_ecog)) + 
  geom_point(alpha = .3) + 
  geom_line(colour = "blue") + 
  theme(legend.position = "bottom") +
  facet_grid(~ph_ecog) +
  labs(title = "Cumulative Probability Adjusted for Ecog performance score", x = "Number of Days", y = "Cumulative Probability of subject's death")

```
Summary: according to the observed value, the subjects' with Ecog performance score = 2 have a comparatively higher chances of survival compared to others.


